---
title: "Statistical analysis of OCT data"
subtitle: "Investigation of Structural Alterations in Inherited Retinal Diseases: A Quantitative SD-OCT-Analysis of Retinal Layer Thicknesses in Light of Underlying Genetic Mutations"
author: "Julien Delarocque"
date: "2023-01-31"
bibliography: sa_oct.bib
link-citations: true
output:
  html_document:
    toc: false
    number_sections: true
    fig_caption: true
    theme: cerulean
    keep_md: true
---

[![DOI](https://zenodo.org/badge/595621353.svg)](https://zenodo.org/badge/latestdoi/595621353)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

# Summary

This script reproduces the essential parts of the analysis presented in [@Gersch2022] on a
mock data set. Consequently, **the results are purely random!**

While the general rationale of the analysis was validated by peer-review, please note
that the script itself was not reviewed. The script is provided under an
[Apache license 2.0](https://www.apache.org/licenses/LICENSE-2.0.txt).

Please cite as [@Delarocque2023].

# Dataset

The here provided data set was randomly generated based on the case numbers, means and
standard deviations by disease entity, layer and segment in the original paper. For
details about layer and segment definition please see [@Gersch2022].

```{r data_import}
library(tidyverse)

dat <- openxlsx::read.xlsx("sim_OCT_data.xlsx") %>%
  mutate(pat_ID = factor(pat_ID),
         sex = factor(sex),
         disease = factor(disease, levels = c("NON", "RP", "STGD", "CRD")),
         eye = factor(eye),
         layer = factor(layer, levels = c("RNFL", "GCL", "IPL", "INL", "OPL", "ONL",
                                          "PR1_2", "RPE", "IRL", "ORL", "Retina")))
```

# Interocular symmetry

Multivariate assessment of interocular symmetry by PCA and euclidean distance.

```{r lat_pca}
# The wide data frame will be used for the multivariate analysis
datw <- dat %>%
  # The following layers summarize other layers, thus need to be removed to avoid
  # cumulative effects of certain layers
  filter(! layer %in% c("Retina","IRL", "ORL")) %>%
  pivot_wider(names_from=layer, values_from=C0:T2, names_glue = "{layer}_{.value}", names_sort = T)

pca_data <- datw %>%
  select(pat_ID, disease, eye, 7:78) %>%
  unite("ID",pat_ID, eye) %>%
  column_to_rownames("ID") %>%
  mutate(across(where(is.numeric), scale))

# Impute missing data for multivariate analysis only
# (in the mock dataset no values are missing)
# pca_data %>% Amelia::missmap()
# pca_data[,-1] <- impute::impute.knn(as.matrix(pca_data[,-1]))$data

source("utils.R")
lat_p1 <- mpca(pca_data[,-1], group=pca_data[,1], scaling = F,
               link=sub("_L|_R","",rownames(pca_data)), plot = F) +
  scale_y_continuous(labels=signs::signs_format()) +
  scale_x_continuous(labels=signs::signs_format())
```

```{r lat_dist, fig.asp=0.5, fig.cap="(A) Principal component analysis of the OCT data. Eye pairs a connected by a line.\n(B) Euclidean distance between eye pairs by group.", dpi=300}
euc_data <- datw %>%
  select(pat_ID, disease) %>%
  distinct() %>%
  left_join(pca_data %>%
              select(-disease) %>%
              rownames_to_column("rn") %>%
              separate(rn, c("pat_ID", "eye"), sep = "_") %>%
              finddists("pat_ID", "eye", scale=F) %>%
              {data.frame(pat_ID = names(.), euclidean = ., row.names = NULL)})

euc_n <- with(euc_data, table(disease))
euc_m1 <- aov(euclidean~disease,euc_data)
# car::leveneTest(euc_m1)
# shapiro.test(resid(euc_m1))
# car::qqPlot(resid(euc_m1))
euc_posthoc <- emmeans::emmeans(euc_m1, dunnett~disease)$contrast %>% as.data.frame()

library(ggpubr)
lat_p2 <- euc_data %>%
  ggplot(aes(disease, euclidean, color=disease)) +
  ggforce::geom_sina(show.legend = F, method="density", seed=90793, maxwidth=1) +
  # adjust to number of wished comparisons
  geom_bracket(xmin="NON", xmax="RP", y.position=14.4, inherit.aes = F,label.size = 2,
               label=sprintf("p = %.3f",euc_posthoc$p.value[euc_posthoc$contrast=="RP - NON"])) +
  scale_x_discrete(labels=function(x) {sprintf("%s\n(n = %d)",x,euc_n[x])}) +
  scale_y_continuous(expand=c(0.05,0.5)) +
  labs(x=NULL, y="Euclidean distance between eyes")

library(patchwork)
lat_p1 + lat_p2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A") &
  ggthemes::scale_color_ptol("Group") &
  theme_light(base_size = 8) &
  theme(panel.grid = element_blank(), legend.position = "bottom", legend.margin = margin(t = -10))
```

```{r lat_dist_aov, results='asis'}
report::report(anova(euc_m1))
```

```{r lat_dist_posthoc}
euc_posthoc %>% knitr::kable(caption = "Dunnett pairwise comparisons among groups. The different group sizes should be accounted for in the interpretation.")
```

# Alterations of the retinal structure in IRDs

## Retinitis pigmentosa

```{r match_rp, cache=TRUE, fig.asp=0.92, fig.cap="**Mean difference in thickness between RP and controls by layer and segment.** Significant differences are denoted with a star (*)."}
datl <- dat %>%
  pivot_longer(C0:T2, names_to = "segment", values_to = "thickness")

matched_rp <- dat %>%
  filter(disease %in% c("NON","RP")) %>%
  select(pat_ID, disease, age, sex) %>%
  distinct() %>%
  match_patients(id="pat_ID", treatment = "disease", control="NON", check = F) %>%
  select(pat_ID, pair) %>%
  left_join(datl) %>%
  droplevels()

library(table1)
table1(~ age + sex | disease, overall = F,
       data=matched_rp %>% select(pat_ID, age, sex, disease) %>% distinct())

rp_uni_comp <- matched_rp %>%
  group_by(layer, segment) %>%
  nest() %>%
  mutate(fit = map(data, ~nlme::gls(thickness ~ disease, data = .,
                                    weights=nlme::varIdent(form = ~ 1 | disease),
                                    correlation=nlme::corSymm(form = ~ 1 | pat_ID),
                                    na.action = na.omit)),
         tid = map(fit, parameters::parameters)) %>%
  unnest(tid) %>%
  ungroup() %>%
  filter(Parameter=="diseaseRP") %>%
  select(-Parameter) %>%
  mutate(adj.p=p.adjust(p,method = "holm"),
         sig.p=adj.p<.05) %>%
  arrange(adj.p)
# p values were computed using a Wald t-distribution approximation
# table(sapply(rp_uni_comp$fit, \(x) shapiro.test(resid(x))$p.value)<0.05)
# plot(rp_uni_comp$fit[[1]])

rp_uni_comp %>%
  select(layer, segment, Coefficient, SE, CI_low, CI_high, t, df_error, p, adj.p) %>%
  head(20) %>%
  knitr::kable(caption = "Top 20 most significantly different segment thicknesses.", digits = 3)

pvals_by_layer(rp_uni_comp)
```

## Cone-rods dystrophie

```{r match_crd, cache=TRUE, fig.asp=0.92, fig.cap="**Mean difference in thickness between CRD and controls by layer and segment.** Significant differences are denoted with a star (*)."}
matched_crd <- dat %>%
  filter(disease %in% c("NON","CRD")) %>%
  select(pat_ID, disease, age, sex) %>%
  distinct() %>%
  match_patients(id="pat_ID", treatment = "disease", control="NON", check = F) %>%
  select(pat_ID, pair) %>%
  left_join(datl) %>%
  droplevels()

table1(~ age + sex | disease, overall = F,
       data=matched_crd %>% select(pat_ID, age, sex, disease) %>% distinct())

crd_uni_comp <- matched_crd %>%
  group_by(layer, segment) %>%
  nest() %>%
  mutate(fit = map(data, ~nlme::gls(thickness ~ disease, data = .,
                                    weights=nlme::varIdent(form = ~ 1 | disease),
                                    correlation=nlme::corSymm(form = ~ 1 | pat_ID),
                                    na.action = na.omit)),
         tid = map(fit, parameters::parameters)) %>%
  unnest(tid) %>%
  ungroup() %>%
  filter(Parameter=="diseaseCRD") %>%
  select(-Parameter) %>%
  mutate(adj.p=p.adjust(p,method = "holm"),
         sig.p=adj.p<.05) %>%
  arrange(adj.p)

crd_uni_comp %>%
  select(layer, segment, Coefficient, SE, CI_low, CI_high, t, df_error, p, adj.p) %>%
  head(20) %>%
  knitr::kable(caption = "Top 20 most significantly different segment thicknesses.", digits = 3)

pvals_by_layer(crd_uni_comp)
```

## Morbus Stargard

```{r match_stgd, cache=TRUE, fig.asp=0.92, fig.cap="**Mean difference in thickness between STGD and controls by layer and segment.** Significant differences are denoted with a star (*)."}
matched_stgd <- dat %>%
  filter(disease %in% c("NON","STGD")) %>%
  select(pat_ID, disease, age, sex) %>%
  distinct() %>%
  match_patients(id="pat_ID", treatment = "disease", control="NON", check = F) %>%
  select(pat_ID, pair) %>%
  left_join(datl) %>%
  droplevels()

table1(~ age + sex | disease, overall = F,
       data=matched_stgd %>% select(pat_ID, age, sex, disease) %>% distinct())

stgd_uni_comp <- matched_stgd %>%
  group_by(layer, segment) %>%
  nest() %>%
  mutate(fit = map(data, ~nlme::gls(thickness ~ disease, data = .,
                                    weights=nlme::varIdent(form = ~ 1 | disease),
                                    correlation=nlme::corSymm(form = ~ 1 | pat_ID),
                                    na.action = na.omit)),
         tid = map(fit, parameters::parameters)) %>%
  unnest(tid) %>%
  ungroup() %>%
  filter(Parameter=="diseaseSTGD") %>%
  select(-Parameter) %>%
  mutate(adj.p=p.adjust(p,method = "holm"),
         sig.p=adj.p<.05) %>%
  arrange(adj.p)

stgd_uni_comp %>%
  select(layer, segment, Coefficient, SE, CI_low, CI_high, t, df_error, p, adj.p) %>%
  head(20) %>%
  knitr::kable(caption = "Top 20 most significantly different segment thicknesses.", digits = 3)

pvals_by_layer(stgd_uni_comp)
```

# Association between subfield thicknesses and visual function

```{r vis_overview, cache=TRUE}
vis_dat <- datl %>%
  filter(!is.na(logMAR)) %>%
  select(pat_ID, logMAR, age, sex, disease) %>%
  distinct()

table1(~logMAR + age + sex | disease, overall = F, data=vis_dat)

vis_n <- with(vis_dat, table(disease))
(vis_m1 <- kruskal.test(vis_dat$logMAR, vis_dat$disease))
vis_ph <- PMCMRplus::kwManyOneDunnTest(logMAR~disease,vis_dat) %>% summary()

vis_dat %>%
  ggplot(aes(disease, logMAR, color=disease)) +
  ggforce::geom_sina(show.legend = F, method="density", seed=1362, maxwidth=.7) +
  geom_bracket(xmin="NON", xmax="STGD", y.position=3.28, inherit.aes = F,label.size = 2,
               label=sprintf("p = %.3f",vis_ph$p.value["STGD",])) +
  geom_bracket(xmin="NON", xmax="CRD", y.position=3.5, inherit.aes = F,label.size = 2,
             label=sprintf("p = %.3f",vis_ph$p.value["CRD",])) +
  ggthemes::scale_color_ptol() +
  labs(x=NULL, y="logMAR (higher values indicate worse vision)") +
  scale_x_discrete(labels=function(x) {sprintf("%s\n(n = %d)",x,vis_n[x])}) +
  theme_light(base_size = 8) +
  theme(panel.grid = element_blank())
```

```{r cor_vis, cache=TRUE, fig.asp=.7, fig.cap="**Relationship between vision as logMAR and segment thicknesses within layers.** The regression line is obtained from a robust mixed model."}
library(robustlmm)
vis_uni_comp <- datl %>%
  filter(!is.na(logMAR)) %>%
  group_by(layer, segment) %>%
  nest() %>%
  mutate(fit = map(data, ~rlmer(logMAR ~ thickness + (1|pat_ID), data=.)),
         tid = map(fit, parameters::parameters)) %>%
  unnest(tid) %>%
  ungroup() %>%
  filter(Parameter=="thickness") %>%
  mutate(adj.p=p.adjust(p,"holm"),
         sig.p=adj.p<.05) %>%
  arrange(adj.p)

vis_uni_comp %>%
  select(layer, segment, Coefficient, SE, CI_low, CI_high, t, p, adj.p) %>%
  head(20) %>%
  knitr::kable(caption = "Top 20 segments most significantly associated with logMAR.", digits = 3)

vis_preds <- vis_uni_comp %>%
  head(20) %>%
  select(layer, segment, data, fit, adj.p) %>%
  mutate(pred = map(fit, ~ggeffects::ggpredict(.)$thickness)) %>%
  unnest(pred)

vis_uni_comp %>%
  head(20) %>%
  select(layer, segment, data, fit, adj.p) %>%
  unnest(data) %>%
  ggplot(aes(thickness, logMAR, color=disease)) +
  geom_point() +
  geom_ribbon(data=vis_preds, mapping=aes(x=x, ymin=conf.low, ymax=conf.high),
              inherit.aes = F, alpha=.2) +
  geom_line(data=vis_preds, mapping=aes(x=x, y=predicted), inherit.aes = F) +
  labs(x="Thickness (µm)") +
  ggthemes::scale_color_ptol("Group") +
  guides(color=guide_legend(nrow = 1, title.hjust = .5)) +
  facet_wrap(~interaction(segment, layer), scales = "free",
             labeller = as_labeller(\(x) {sub("_","/",sub("([A-Z][0-2])\\.(.+)","\\2 \\1",x))})) +
  theme_light(base_size = 8) +
  theme(panel.grid = element_blank(),
        legend.position = "top",
        legend.margin = margin(r=10, b = -4),
        legend.direction = "horizontal")
```

# Licence

Copyright 2023 Julien Delarocque
 
Licensed under the Apache License, Version 2.0 (the "License"); you may not use this file
except in compliance with the License. You may obtain a copy of the License at
http://www.apache.org/licenses/LICENSE-2.0
 
Unless required by applicable law or agreed to in writing, software distributed under the
License is distributed on an "AS IS" BASIS, WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND,
either express or implied. See the License for the specific language governing
permissions and limitations under the License.

# References

::: {#refs}
:::

# Session information

```{r session_info, results='asis'}
pander::pander(sessionInfo())
```
